Comparison of FDR methods on 16S microbiome data with SummarizedBenchmark

### 2018-02-19 datasets/microbiome/microbiome-schubert.Rmd
Start splitting microbiome analyses

---
title: "Comparison of FDR methods on 16S microbiome data with SummarizedBenchmark"
author: "Claire Duvallet"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = "/n/irizarryfs01_backed_up/kkorthauer/FDR")
```

### Summary

@@

### Workspace setup

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
# project directory & data/results folders
setwd("/Users/claire/github/benchmark-fdr/datasets/microbiome/")
datdir <- "/Users/claire/github/benchmark-fdr/datasets/microbiome/DATA/"
resdir <- "/Users/claire/github/benchmark-fdr/datasets/microbiome/RESULTS/"

# wrangling & plotting tools
library(data.table)
#library(readxl)
library(dplyr)
library(ggplot2)
library(magrittr)
library(R.utils)
library(cowplot)

# benchmark methods
library(IHW)
library(ashr)
library(qvalue)
library(swfdr)
library(fdrtool)
library(FDRreg)

# comparison tools/functions
library(SummarizedBenchmark)  
sourceDirectory("/Users/claire/github/benchmark-fdr/datasets/R/")

## set up parallel backend 
library(BiocParallel)
cores <- as.numeric(Sys.getenv("SLURM_NTASKS"))
multicoreParam <- MulticoreParam(workers = cores)
```

Let's just keep going and hopefully it's okay...

### CDI Schubert - diarrhea dataset

Here we download and analyze the CDI Schubert diarrhea (C. diff and non-C.diff diarrhea) dataset. 
We'll download the processed OTU tables from Zenodo and unzip them in the 
`benchmark-fdr/datasets/microbiome/DATA` folder.

#### Data download

```{bash, schubert-download, eval=FALSE}
cd /Users/claire/github/benchmark-fdr/datasets/microbiome/DATA/
curl -O "https://zenodo.org/record/840333/files/cdi_schubert_results.tar.gz"
tar -xzvf cdi_schubert_results.tar.gz
cd ..
```

@@

```{r, schubert-readdata, eval=TRUE}
otu_path = paste0(datdir, "cdi_schubert_results/RDP/cdi_schubert.otu_table.100.denovo.rdp_assigned")
meta_path = paste0(datdir, "cdi_schubert_results/cdi_schubert.metadata.txt")

# DiseaseState labels to keep
labels <- c("H", "nonCDI", "CDI")

# Metadata column that has sample IDs
col_label <- 'sample_id'

# load OTU table and metadata
otu <- read.table(otu_path)
meta <- read.csv(meta_path, sep='\t')

# Keep only samples with the right DiseaseState metadata
meta <- meta %>% filter(DiseaseState %in% labels)

# Keep only samples with both metadata and 16S data
keep_samples <- intersect(colnames(otu), meta[, 1])
otu <- otu[, keep_samples]
meta <- meta %>% filter(get(col_label) %in% keep_samples)

```

@@

```{r, schubert-cleandata, eval=TRUE}
remove_shallow_smpls <- function(df, n_reads) {
  # Removes samples with fewer than n_reads from dataframe df.
  # df has OTUs in rows and samples in columns
  return(df[, colSums(df) > n_reads])
  }

remove_shallow_otus <- function(df, n_reads){
  return(df[rowSums(df) > n_reads, ])
  }

## Remove OTUs with fewer than 10 reads
otu <- remove_shallow_otus(otu, 10)

# Remove samples with fewer than sample_Reads reads
sample_reads <- 100
otu <- remove_shallow_smpls(otu, sample_reads)

# Update metadata with new samples
meta <- meta %>% filter(get(col_label) %in% colnames(otu))

# Remove empty OTUs
otu <- otu[rowSums(otu) > 0, ]

# Convert to relative abundance
abun_otu <- t(t(otu) / rowSums(t(otu)))

# Add pseudo counts
#minabun <- min(abun_otu[abun_otu > 0]) # Use this to figure out what pseudo-count value to add
zeroabun <- 0
abun_otu <- abun_otu + zeroabun

```

@@

```{r, schubert-pvals, eval=TRUE}
resfile <- paste0(resdir, "schubert_results_",
                  nrow(abun_otu), "_OTUs.RData")
if (!file.exists(resfile)){
  # Get case and control indices
  case_idx <- meta$DiseaseState %in% c("CDI", "nonCDI")
  ctrl_idx <- meta$DiseaseState %in% c("H")

  # Calculate pvalues, effects, and stderr
  pvals <- c()
  teststats <- c()
  ses <- c()
  effs <- c()
  mean_abuns <- c()
  mean_abuns_present <- c()
  ubis <- c()

  casedf <- abun_otu[, case_idx]
  ctrldf <- abun_otu[, ctrl_idx]
  for (o in rownames(abun_otu)) {
    # wilcoxon p value  
    w <- wilcox.test(casedf[o, ], 
                     ctrldf[o, ])
    p <- w$p.value
    teststat <- w$statistic

    pvals <- c(pvals, p)
    teststats <- c(teststats, teststat)

    # standard error of the OTU abundance, across all samples
    ses <- c(ses, 
             sd(abun_otu[o, ])/sqrt(length(abun_otu[o, ]))
             )

    # mean OTU abundance across all samples (after removing pseudo-count)
    mean_abuns <- c(mean_abuns, 
                    mean(abun_otu[o, ] - zeroabun)
                    )

    # mean OTU abundance across only samples with the OTU present
    mean_abuns_present <- c(mean_abuns_present, 
                            sum(abun_otu[o, ] - zeroabun) / sum(abun_otu[o, ] > zeroabun)
                            )

    # ubiquity of OTU across all samples
    ubis <- c(ubis, 
              sum(abun_otu[o, ] > zeroabun) / length(abun_otu[o, ])
              )

    # effect (logfold difference)
    effs <- c(effs, 
              log(mean(abun_otu[o, ctrl_idx])/mean(abun_otu[o, case_idx]))
              )

  }

  res <- data.frame(otu = rownames(abun_otu), 
                    pval = pvals, wilcox_teststat = teststats,
                    SE = ses, effect_size = effs, 
                    mean_abun = mean_abuns, mean_abun_present = mean_abuns_present,
                    ubiquity = ubis)
  res <- res %>%
    mutate(test_statistic = qnorm(exp(log(pval)-log(2)), lower.tail=FALSE) * sign(effect_size))

  save(res, file=resfile)

}else{
  load(resfile)
}
head(res)
```

@@

```{r, schubert-addphylo, eval=TRUE}
res <- res %>% separate(otu, 
                        c("kingdom", "phylum", "class", "order", "family", "genus", "species", "denovo"), 
                        sep=";", remove = FALSE)

```

#### Check Covariate Diagnostics

@@

##### Ubiquity

@@

##### Mean abundance (across non-zero samples)

@@

##### Phylogeny

@@

```{r, schubert-phylo, fig.width=10, fig.height=6, eval=FALSE}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

Let's use `ubiquity` as our `ind_covariate`

```{r, schubert-labelcovariate, eval=TRUE}
res <- res %>% mutate(ind_covariate = ubiquity)
```

Also (hacky), remove test_statistic so that ash doesn't run.

```{r, schubert-rm-ash, eval=TRUE}
res <- res %>% 
  mutate(test_statistic = NA) %>%
  mutate(effect_size = NA)
```

#### Set up BenchDesign object

@@

```{r, schubert-sb, results="hide", message=FALSE, eval=TRUE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

  resfile <- paste0(resdir, "schubert_summarizedBenchmark_",
                    nrow(res), "_nmids_", nmids, ".RData")
  duration <- NA
  if (!file.exists(resfile)){
    t1 <- proc.time()
    sb <- bd %>% buildBench(data=res, ftCols = "ind_covariate")
                            #parallel=TRUE, BPPARAM=multicoreParam)
    metadata(sb)$data_download_link <- "https://zenodo.org/record/840333/files/cdi_schubert_results.tar.gz"
    save(sb, file=resfile)
    duration <- round((proc.time()-t1)[3]/60,1)
  }else{
    load(resfile)
  }
}

```

Just for future runs, where I load the file instead of re-calculating it, here are the errors I got:

```
[1] 5
Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !NaNs produced<simpleError in if (sum(D^2) > qchisq(0.9, nmids - 2 - df)) {    warning(paste0("f(z) misfit = ", round(D, 1), ".  Rerun with increased df."))}: missing value where TRUE/FALSE needed>

[1] 50
Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !f(z) misfit = -21.6.  Rerun with increased df.f(z) misfit = -22.5.  Rerun with increased df.f(z) misfit = -10.  Rerun with increased df.f(z) misfit = 32.  Rerun with increased df.f(z) misfit = 14.  Rerun with increased df.f(z) misfit = -32.2.  Rerun with increased df.f(z) misfit = 6.9.  Rerun with increased df.f(z) misfit = -1.3.  Rerun with increased df.f(z) misfit = 0.7.  Rerun with increased df.f(z) misfit = 3.4.  Rerun with increased df.f(z) misfit = 5.5.  Rerun with increased df.f(z) misfit = 0.1.  Rerun with increased df.f(z) misfit = -0.9.  Rerun with increased df.f(z) misfit = -1.5.  Rerun with increased df.f(z) misfit = -1.9.  Rerun with increased df.f(z) misfit = -0.8.  Rerun with increased df.f(z) misfit = -2.4.  Rerun with increased df.f(z) misfit = 1.7.  Rerun with increased df.f(z) misfit = 0.5.  Rerun with increased df.f(z) misfit = 2.  Rerun with increased df.f(z) misfit = 0.1.  Rerun with increased df.f(z) misfit = 1.4.  Rerun with increased df.f(z) misfit = 0.2.  Reru... <truncated>

[1] 150
Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !Censored sample for null model estimation has only size 0 !f(z) misfit = 83.4.  Rerun with increased df.f(z) misfit = -8.5.  Rerun with increased df.f(z) misfit = -13.6.  Rerun with increased df.f(z) misfit = -17.1.  Rerun with increased df.f(z) misfit = -6.4.  Rerun with increased df.f(z) misfit = -19.5.  Rerun with increased df.f(z) misfit = -17.8.  Rerun with increased df.f(z) misfit = -3.  Rerun with increased df.f(z) misfit = -21.1.  Rerun with increased df.f(z) misfit = 28.6.  Rerun with increased df.f(z) misfit = -26.  Rerun with increased df.f(z) misfit = 59.3.  Rerun with increased df.f(z) misfit = -17.1.  Rerun with increased df.f(z) misfit = 12.1.  Rerun with increased df.f(z) misfit = -28.3.  Rerun with increased df.f(z) misfit = -27.6.  Rerun with increased df.f(z) misfit = 83.7.  Rerun with increased df.f(z) misfit = -10.5.  Rerun with increased df.f(z) misfit = -22.6.  Rerun with increased df.f(z) misfit = -23.3.  Rerun with increased df.f(z) misfit = 37.3.  Rerun with increased df.f(z) misfit = -18.  Rerun with increased df.f(z... <truncated>>
```

```{r, echo=FALSE, eval=FALSE}
if(!is.na(duration)){
  message("This step took ", duration, " minutes for ", nrow(res), " OTUs using ",
        cores, " cores.")
}
```

@@

```{r, schubert-metrics, eval=TRUE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

@@

Debugging note: to look at the matrix of qvalues, run `assays(sb)[["qvalue"]]`

```{r, schubert-plot, width=15, height=15, eval=TRUE}

# plot nrejects by method overall and stratified by covariate
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )

```

@@

```{r, schubert-genus-covariates}
methods <- c( "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf" )
plotCovariateBoxplots( sb, alpha=0.1, nsets=6, methods=methods)
```

@@

And we'll clean up the workspace before moving on to the next dataset.

```{r, schubert-cleanup, eval=FALSE}
rm(res)
rm(bd)
rm(sb)
rm(pf)
```

### OB Goodrich - obesity dataset

Let's repeat these analyses for the OB Goodrich dataset.

Here we download and analyze the CDI Schubert diarrhea (C. diff and non-C.diff diarrhea) dataset. 
We'll download the processed OTU tables from Zenodo and unzip them in the 
`benchmark-fdr/datasets/microbiome/DATA` folder.

#### Data download

```{bash, goodrich-download, eval=FALSE}
cd /Users/claire/github/benchmark-fdr/datasets/microbiome/DATA/
curl -O "https://zenodo.org/record/840333/files/ob_goodrich_results.tar.gz"
tar -xzvf ob_goodrich_results.tar.gz
cd ..
```

```{r, goodrich-readdata}
otu_path = paste0(datdir, "ob_goodrich_results/RDP/ob_goodrich.otu_table.100.denovo.rdp_assigned")
meta_path = paste0(datdir, "ob_goodrich_results/ob_goodrich.metadata.txt")

# DiseaseState labels to keep
labels <- c("H", "OB")

# Metadata column that has sample IDs
col_label <- "X"

# load OTU table and metadata
otu <- read.table(otu_path)
meta <- read.csv(meta_path, sep='\t')

# Keep only samples with the right DiseaseState metadata
meta <- meta %>% filter(DiseaseState %in% labels)

# Keep only samples with both metadata and 16S data
keep_samples <- intersect(colnames(otu), meta[, 1])
otu <- otu[, keep_samples]
meta <- meta %>% filter(get(col_label) %in% keep_samples)

```

```{r, goodrich-cleandata}
remove_shallow_smpls <- function(df, n_reads) {
  # Removes samples with fewer than n_reads from dataframe df.
  # df has OTUs in rows and samples in columns
  return(df[, colSums(df) > n_reads])
}

remove_shallow_otus <- function(df, n_reads){
  return(df[rowSums(df) > n_reads, ])
}

remove_rare_otus <- function(df, perc_samples){
  return(df[rowSums(df > 0) / dim(df)[2] > perc_samples, ])
}

## Remove OTUs with fewer than 10 reads
otu <- remove_shallow_otus(otu, 10)
print(dim(otu))
## Remove OTUs in fewer than 1% of samples
otu <- remove_rare_otus(otu, 0.01)
print(dim(otu))

# Remove samples with fewer than sample_reads reads
sample_reads <- 100
otu <- remove_shallow_smpls(otu, sample_reads)

# Update metadata with new samples
meta <- meta %>% filter(get(col_label) %in% colnames(otu))

# Remove empty OTUs
otu <- otu[rowSums(otu) > 0, ]

# Convert to relative abundance
abun_otu <- t(t(otu) / rowSums(t(otu)))

# Add pseudo counts
#minabun <- min(abun_otu[abun_otu > 0]) # Use this to figure out what pseudo-count value to add
zeroabun <- 0
abun_otu <- abun_otu + zeroabun

```

Next, we need to calculate the pvalues, effect size, and standard error for each OTU.
Here, we'll compare lean vs. obese. We'll put these results into a dataframe, and label 
the columns with the standardized names for downstream use 
(`pval`, `SE`, `effect_size`, `test_statistic`). The test statistic is the one returned 
by `wilcox.test()`.

Note that OB Goodrich has ~70,000 OTUs, so this is going to take a while. There are probably
faster ways to write this, but I don't know them.

```{r}
dim(abun_otu)
```

```{r, goodrich-pvals}
resfile <- paste0(resdir, "goodrich_results_",
                  nrow(abun_otu), "_OTUs.RData")
if (!file.exists(resfile)){
  # Get case and control indices
  case_idx <- meta$DiseaseState %in% c("OB")
  ctrl_idx <- meta$DiseaseState %in% c("H")

  # Calculate pvalues, effects, and stderr
  pvals <- c()
  teststats <- c()
  ses <- c()
  effs <- c()
  mean_abuns <- c()
  mean_abuns_present <- c()
  ubis <- c()

  casedf <- abun_otu[, case_idx]
  ctrldf <- abun_otu[, ctrl_idx]
  for (o in rownames(abun_otu)) {
    # wilcoxon p value  
    w <- wilcox.test(casedf[o, ], 
                     ctrldf[o, ])
    p <- w$p.value
    teststat <- w$statistic

    pvals <- c(pvals, p)
    teststats <- c(teststats, teststat)

    # standard error of the OTU abundance, across all samples
    ses <- c(ses, 
             sd(abun_otu[o, ])/sqrt(length(abun_otu[o, ]))
             )

    # mean OTU abundance across all samples (after removing pseudo-count)
    mean_abuns <- c(mean_abuns, 
                    mean(abun_otu[o, ] - zeroabun)
                    )

    # mean OTU abundance across only samples with the OTU present
    mean_abuns_present <- c(mean_abuns_present, 
                            sum(abun_otu[o, ] - zeroabun) / sum(abun_otu[o, ] > zeroabun)
                            )

    # ubiquity of OTU across all samples
    ubis <- c(ubis, 
              sum(abun_otu[o, ] > zeroabun) / length(abun_otu[o, ])
              )

    # effect (logfold difference)
    effs <- c(effs, 
              log(mean(abun_otu[o, ctrl_idx])/mean(abun_otu[o, case_idx]))
              )

  }

  res <- data.frame(otu = rownames(abun_otu), 
                    pval = pvals, test_statistic = teststats,
                    SE = ses, effect_size = effs, 
                    mean_abun = mean_abuns, mean_abun_present = mean_abuns_present,
                    ubiquity = ubis)
  res <- res %>%
    mutate(test_statistic = qnorm(exp(log(pval)-log(2)), lower.tail=FALSE) * sign(effect_size))

  save(res, file=resfile)

}else{
  load(resfile)
}
head(res)
```

```{r, goodrich-addphylo}
res <- res %>% separate(otu, 
                        c("kingdom", "phylum", "class", "order", "family", "genus", "species", "denovo"), 
                        sep=";", remove = FALSE)

```

#### Check Covariate Diagnostics

##### Ubiquity

```{r, goodrich-ubi, fig.width=10, fig.height=3.5}
strat_hist(res, pvalue="pval", covariate="ubiquity", maxy=5, binwidth=0.05, numQ=4)
rank_scatter(res, pvalue="pval", covariate="ubiquity")
```

##### Mean abundance (across non-zero samples)

```{r, goodrich-abun, fig.width=10, fig.height=3.5}
strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=5, binwidth=0.05, numQ=4)
rank_scatter(res, pvalue="pval", covariate="mean_abun_present")
```

##### Phylogeny

```{r, goodrich-phylo, fig.width=10, fig.height=6, eval=FALSE}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

Let's use `ubiquity` as our `ind_covariate`

```{r, goodrich-labelcovariate}
res <- res %>% mutate(ind_covariate = ubiquity)
```

Also (hacky), remove test_statistic so that ash doesn't run.

```{r, goodrich-rm-ash}
res <- res %>% 
  mutate(test_statistic = NA) %>%
  mutate(effect_size = NA)
```

#### Set up BenchDesign object

First, we'll create an object of `BenchDesign` class to hold the data and 
add the benchmark methods to the `BenchDesign` object.

Next, we'll construct the `SummarizedBenchmark` object, which will run
the functions specified in each method (these are actually sourced in from the
helper scripts).

```{r, goodrich-sb, results="hide", message=FALSE, eval=TRUE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

  resfile <- paste0(resdir, "goodrich_summarizedBenchmark_",
                    nrow(res), "_nmids_", nmids, ".RData")
  duration <- NA
  if (!file.exists(resfile)){
    t1 <- proc.time()
    sb <- bd %>% buildBench(data=res, ftCols = "ind_covariate")
                            #parallel=TRUE, BPPARAM=multicoreParam)
    metadata(sb)$data_download_link <- "https://zenodo.org/record/840333/files/cdi_schubert_results.tar.gz"
    save(sb, file=resfile)
    duration <- round((proc.time()-t1)[3]/60,1)
  }else{
    load(resfile)
  }
}
```

The outputs of this call are:

```
[1] 5
Squarem-1
Objective fn: 715262
Objective fn: 104929  Extrapolation: 0  Steplength: 1
Objective fn: 95918.9  Extrapolation: 1  Steplength: 2.44204
Objective fn: 95614.3  Extrapolation: 1  Steplength: 2.72536
Objective fn: 95583.3  Extrapolation: 1  Steplength: 2.8568
Objective fn: 95578.1  Extrapolation: 1  Steplength: 3.54373
Objective fn: 95576.3  Extrapolation: 1  Steplength: 3.52004
Objective fn: 95575.4  Extrapolation: 1  Steplength: 4
.
.
.
NaNs produced<simpleError in if (sum(D^2) > qchisq(0.9, nmids - 2 - df)) {    warning(paste0("f(z) misfit = ", round(D, 1), ".  Rerun with increased df."))}: missing value where TRUE/FALSE needed>

[1] 50
Squarem-1
Objective fn: 715262
Objective fn: 104929  Extrapolation: 0  Steplength: 1
Objective fn: 95918.9  Extrapolation: 1  Steplength: 2.44204
Objective fn: 95614.3  Extrapolation: 1  Steplength: 2.72536
Objective fn: 95583.3  Extrapolation: 1  Steplength: 2.8568
Objective fn: 95578.1  Extrapolation: 1  Steplength: 3.54373
Objective fn: 95576.3  Extrapolation: 1  Steplength: 3.52004
Objective fn: 95575.4  Extrapolation: 1  Steplength: 4
Objective fn: 95574.9  Extrapolation: 1  Steplength: 3.36769
Objective fn: 95574.5  Extrapolation: 1  Steplength: 5.09115
Objective fn: 95574.3  Extrapolation: 1  Steplength: 2.52066
Objective fn: 95574  Extrapolation: 1  Steplength: 13.3966
.
.
.
Objective fn: 95572.9  Extrapolation: 1  Steplength: 1.61619
f(z) misfit = -8.7.  Rerun with increased df.f(z) misfit = 33.3.  Rerun with increased df.f(z) misfit = -9.4.  Rerun with increased df.f(z) misfit = -37.3.  Rerun with increased df.f(z) misfit = 49.2.  Rerun with increased df.f(z) misfit = -36.9.  Rerun with increased df.f(z) misfit = 10.6.  Rerun with increased df.f(z) misfit = 13.3.  Rerun with increased df.f(z) misfit = -18.3.  Rerun with increased df.f(z) misfit = 20.1.  Rerun with increased df.f(z) misfit = -21.6.  Rerun with increased df.f(z) misfit = 23.4.  Rerun with increased df.f(z) misfit = -27.  Rerun with increased df.f(z) misfit = 17.2.  Rerun with increased df.f(z) misfit = -8.2.  Rerun with increased df.f(z) misfit = 4.3.  Rerun with increased df.f(z) misfit = 3.4.  Rerun with increased df.f(z) misfit = 2.4.  Rerun with increased df.f(z) misfit = -2.3.  Rerun with increased df.f(z) misfit = -10.7.  Rerun with increased df.f(z) misfit = -2.4.  Rerun with increased df.f(z) misfit = 15.4.  Rerun with increased df.f(z) misf... <truncated>

[1] 150
Squarem-1
Objective fn: 715262
Objective fn: 104929  Extrapolation: 0  Steplength: 1
Objective fn: 95918.9  Extrapolation: 1  Steplength: 2.44204
Objective fn: 95614.3  Extrapolation: 1  Steplength: 2.72536
Objective fn: 95583.3  Extrapolation: 1  Steplength: 2.8568
Objective fn: 95578.1  Extrapolation: 1  Steplength: 3.54373
Objective fn: 95576.3  Extrapolation: 1  Steplength: 3.52004
Objective fn: 95575.4  Extrapolation: 1  Steplength: 4
Objective fn: 95574.9  Extrapolation: 1  Steplength: 3.36769
Objective fn: 95574.5  Extrapolation: 1  Steplength: 5.09115
Objective fn: 95574.3  Extrapolation: 1  Steplength: 2.52066
Objective fn: 95574  Extrapolation: 1  Steplength: 13.3966
.
.
.
f(z) misfit = -20.2.  Rerun with increased df.f(z) misfit = 13.6.  Rerun with increased df.f(z) misfit = -3.8.  Rerun with increased df.f(z) misfit = 9.  Rerun with increased df.f(z) misfit = -21.2.  Rerun with increased df.f(z) misfit = 28.8.  Rerun with increased df.f(z) misfit = 41.9.  Rerun with increased df.f(z) misfit = -13.2.  Rerun with increased df.f(z) misfit = -12.8.  Rerun with increased df.f(z) misfit = 15.2.  Rerun with increased df.f(z) misfit = -18.1.  Rerun with increased df.f(z) misfit = -19.1.  Rerun with increased df.f(z) misfit = -18.6.  Rerun with increased df.f(z) misfit = -26.  Rerun with increased df.f(z) misfit = -7.2.  Rerun with increased df.f(z) misfit = 67.1.  Rerun with increased df.f(z) misfit = 27.2.  Rerun with increased df.f(z) misfit = -16.4.  Rerun with increased df.f(z) misfit = -21.1.  Rerun with increased df.f(z) misfit = -25.9.  Rerun with increased df.f(z) misfit = -7.5.  Rerun with increased df.f(z) misfit = 10.1.  Rerun with increased df.f(z)... <truncated>
```

```{r, echo=FALSE, eval=FALSE}
if(!is.na(duration)){
  message("This step took ", duration, " minutes for ", nrow(res), " OTUs using ",
        cores, " cores.")
}
```

```{r, goodrich-metrics, eval=TRUE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

Debugging note: to look at the matrix of qvalues, run `assays(sb)[["qvalue"]]`

```{r, goodrich-plot, width=15, height=15, eval=TRUE}
# plot nrejects by method overall and stratified by covariate
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
#plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE )
```

There is no upsetR plot for OB Goodrich for now because only lfdr is returning any rejections.

```{r, eval=FALSE}
assays(sb)[["qvalue"]][, "ashs"] %>% max()
sum(is.na(assays(sb)[["qvalue"]][, "scott-empirical"]))
sum(is.na(assays(sb)[["qvalue"]][, "lfdr"]))

```

## Goodrich, genus-level analysis

```{r, goodrich-genus}
## Collapse baxter OTU table to genus level

# Add column with otu names
genus_df <- as.data.frame(abun_otu)
genus_df$otu <- rownames(genus_df)
# Melt into tidy format
genus_df <- genus_df %>% melt(id.var="otu", variable.name="sample", value.name="abun")
# Split into separate taxonomic levels
genus_df <- genus_df %>% 
  separate(otu, c("kingdom", "phylum", "class", "order", "family", "genus", "species", "denovo"), 
           sep=";", remove = FALSE)

# Get rid of unannoated genera, and sum abundances for genera
genus_df <- genus_df %>% 
  filter(genus != "g__") %>% 
  group_by(genus, sample) %>% 
  summarise(total_abun = sum(abun))

# Convert back to longform
genus_df <- genus_df %>% spread(genus, total_abun) %>% as.data.frame
rownames(genus_df) <- genus_df$sample
genus_df <- genus_df %>% select(-sample) %>% t

# And re-order columns to match metadata
genus_df <- genus_df[, match(meta$X, colnames(genus_df))]
```

```{r, goodrich-genus-pvals}
resfile <- paste0(resdir, "goodrich_results_",
                  nrow(genus_df), "_genera.RData")
if (!file.exists(resfile)){
  # Get case and control indices
  case_idx <- meta$DiseaseState %in% c("OB")
  ctrl_idx <- meta$DiseaseState %in% c("H")

  # Calculate pvalues, effects, and stderr
  pvals <- c()
  teststats <- c()
  ses <- c()
  effs <- c()
  mean_abuns <- c()
  mean_abuns_present <- c()
  ubis <- c()

  casedf <- genus_df[, case_idx]
  ctrldf <- genus_df[, ctrl_idx]
  for (o in rownames(genus_df)) {
    # wilcoxon p value  
    w <- wilcox.test(casedf[o, ], 
                     ctrldf[o, ])
    p <- w$p.value
    teststat <- w$statistic

    pvals <- c(pvals, p)
    teststats <- c(teststats, teststat)

    # standard error of the OTU abundance, across all samples
    ses <- c(ses, 
             sd(genus_df[o, ])/sqrt(length(genus_df[o, ]))
             )

    # mean OTU abundance across all samples (after removing pseudo-count)
    mean_abuns <- c(mean_abuns, 
                    mean(genus_df[o, ] - zeroabun)
                    )

    # mean OTU abundance across only samples with the OTU present
    mean_abuns_present <- c(mean_abuns_present, 
                            sum(genus_df[o, ] - zeroabun) / sum(genus_df[o, ] > zeroabun)
                            )

>>>>

---
title: "Comparison of FDR methods on 16S microbiome data with SummarizedBenchmark"
author: "Claire Duvallet"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
----

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

Here we download and analyze the CDI Schubert diarrhea (C. diff and non-C. diff diarrhea) dataset. 
We'll download the processed OTU tables from Zenodo and unzip them in the 
`benchmark-fdr/datasets/microbiome/DATA` folder.

@@

# Workspace setup

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(magrittr)
library(SummarizedBenchmark)

## load helper functions
for (f in list.files("../R", "\\.(r|R)$", full.names = TRUE)) {
    source(f)
}

## project data/results folders
datdir <- "data"
resdir <- "results"
dir.create(datdir, showWarnings = FALSE)
dir.create(resdir, showWarnings = FALSE)
```

# Data Preparation

```{r, schubert-download}
cdi_data <- file.path(datdir, "cdi_schubert_results.tar.gz")
if (!file.exists(cdi_data)) {
    download.file("https://zenodo.org/record/840333/files/cdi_schubert_results.tar.gz",
                  destfile = cdi_data)
}

cdi_dir <- file.path(datdir, "cdi_schubert_results")
if (!file.exists(file.path(cdi_dir, "cdi_schubert.metadata.txt"))) {
    untar(cdi_data, exdir = datdir)
}
```

@@

```{r, schubert-readdata, eval=TRUE}
## load OTU table and metadata
otu <- read.table(file.path(cdi_dir, "RDP",
                            "cdi_schubert.otu_table.100.denovo.rdp_assigned"))
meta <- read.csv(file.path(cdi_dir, "cdi_schubert.metadata.txt"), sep = '\t')

# Keep only samples with the right DiseaseState metadata
meta <- filter(meta, DiseaseState %in% c("H", "nonCDI", "CDI"))

# Keep only samples with both metadata and 16S data
keep_samples <- intersect(colnames(otu), meta$sample_id)
otu <- otu[, keep_samples]
meta <- filter(meta, 'sample_id' %in% keep_samples)
```

@@

```{r, schubert-cleandata, eval=TRUE}
min_otu <- 10
min_sample <- 100

## Remove OTUs w/ <= min reads, samples w/ <= min reads
otu <- otu[, colSums(otu) > min_otu]
otu <- otu[rowSums(otu) > min_sample, ]

## Update metadata with new samples
meta <- filter(meta, sample_id %in% colnames(otu))

## Remove empty OTUs
otu <- otu[rowSums(otu) > 0, ]

## Convert to relative abundance
abun_otu <- t(t(otu) / rowSums(t(otu)))

## Add pseudo counts
zeroabun <- 0
abun_otu <- abun_otu + zeroabun
```

# Data Analysis

## Differential Testing

@@

```{r, schubert-pvals, eval=TRUE}
resfile <- file.path(resdir, paste0("schubert_results_", nrow(abun_otu), "_OTUs.rda"))
if (!file.exists(resfile)) {
    res <- test_microbiome(abundance = abun_otu, shift = zeroabun,
                           is_case = meta$DiseaseState %in% c("CDI", "nonCDI"))
    save(res, file=resfile)
} else {
  load(resfile)
}
head(res)
```

@@

```{r, schubert-addphylo, eval=TRUE}
res <- dplyr::separate(res, otu, 
                       c("kingdom", "phylum", "class", "order",
                         "family", "genus", "species", "denovo"), 
                       sep = ";", remove = FALSE)
```

## Covariate Diagnostics

@@

### Ubiquity

@@

### Mean Abundance (across non-zero samples)

@@

### Phylogeny

@@

```{r, schubert-phylo, fig.width=10, fig.height=6, eval=FALSE}
ggplot(res, aes(x=pval)) +
    geom_histogram() +
    facet_wrap(~phylum, scales = "free")
```

## Multiple-Testing Correction

Let's use `ubiquity` as our `ind_covariate`.

```{r, schubert-labelcovariate, eval=TRUE}
res <- dplyr::mutate(res, ind_covariate = ubiquity)
```

@@

```{r, schubert-sb, results="hide", message=FALSE, eval=TRUE}
for (nmids in c(5, 50, 150)) {
    print(nmids)
    bd <- initializeBenchDesign(nmids = nmids)
    ## remove ASH from comparison
    bd <- dropBMethod(bd, "ashs")

    resfile <- file.path(resdir, paste0("schubert_summarizedBenchmark_",
                                        nrow(res),  "_nmids_", nmids, ".rda"))
    duration <- NA
    if (!file.exists(resfile)) {
        sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
        metadata(sb)$data_download_link <-
                       "https://zenodo.org/record/840333/files/cdi_schubert_results.tar.gz"
        save(sb, file = resfile)
    } else {
        load(resfile)
    }
}
```

## Benchmark Metrics

@@

```{r, schubert-metrics, eval=TRUE}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

@@

```{r, schubert-plot, width=15, height=15, eval=TRUE}
# plot nrejects by method overall and stratified by covariate
rejections_scatter(sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )
```

@@

```{r, schubert-genus-covariates}
methods <- c("lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf")
plotCovariateBoxplots(sb, alpha = 0.1, nsets = 6, methods = methods)
```

@@

# Session Info

```{r}
sessionInfo()
```
